---
title: "Final Project Report: Analysis of Coffee Consumption and Health Metrics"
subtitle: "Inferential Statistics 2025/2026"
author: "Andrea Recchiuti and Rafaelle Richel Pearl"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract
(say from 3-8 lines) describing the problem and what it has been done

# Introduction
motivation about the problem and possibly pointers to the literature

# Data
Data description and exploration: describes where the data comes from and performs some quick exploratory analysis by means of graphs and summaries
```{r}
library(ggplot2)
coffee_data <- read.csv("synthetic_coffee_health_10000.csv", sep = ",")
summary(coffee_data)

# hist for sleep hours
ggplot(coffee_data, aes(x = Sleep_Hours)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Daily Sleep Hours", x = "Hours of Sleep", y = "Count")

# boxplot of caffeine intake per gender
ggplot(coffee_data, aes(x = Gender, y = Caffeine_mg, fill = Gender)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Caffeine Intake Distribution by Gender", y = "Caffeine (mg)")

# barchart health issues
ggplot(coffee_data, aes(x = Health_Issues, fill = Health_Issues)) +
  geom_bar() +
  scale_fill_brewer(palette = "Reds") +
  theme_minimal() +
  labs(title = "Frequency of Reported Health Issues", x = "Severity", y = "Number of Participants")

```

```{r}
# age group histogram
ggplot(coffee_data, aes(x = Age)) +
  geom_bar(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Distribution of Age Groups", x = "Age Group", y = "Number of Participants")
```

# Statistical Methods
statistical methods applied and the results obtained

### Linear Regression
```{r}
model <- lm(Heart_Rate ~ Caffeine_mg + Sleep_Hours + BMI + Physical_Activity_Hours, data = coffee_data)

summary(model)



```

Caffeine / day -> categorical variable (>400, <400)
then count how many in each category

t test on two subgroups and check if there is a difference in mean in heart Heart_Rate

and then do linear regression to find out what variables affect Heart_Rate





# Results
Conclusions